Variational Monte Carlo on Bosonic Systems

### 2018-04-05 project_1/article/article.tex
Spell check non-interacting section

    \subsection{Non-interacting harmonic oscillators}
        We start by looking at a simple system of non-interacting harmonic
        oscillators, where $a = 0$ and $\beta = 1$. This means that there is no
        hard-shell interaction between the particles and that we use a spherical
        potential trap.  We thus get the trial  wavefunction
        \begin{align}
            \Psi_T(\vf{r})
            = \Phi_T(\vf{r})
            = \prod_{i = 1}^N \exp\para{
                -\alpha |\vf{r}_i|^2
            },
        \end{align}
        where $|\vf{r}_i| = r_i$. As $a = 0$ the interaction term,
        $w(\vf{r}_i, \vf{r}_j)$, vanishes and the Hamiltonian is given by
        (in the spherical case)
        \begin{align}
            H &= \sum_{i = 1}^N h(\vf{r}_i)
            = \sum_{i = 1}^N \Biggl(
                -\frac{\hbar^2}{2m}\nabla_i^2
                + \half m \omega^2 r_i^2
            \Biggr).
        \end{align}
        To find the drift force and the local energy we have to compute the
        gradient and the Laplacian of the trial wavefunction. The gradient is
        given by
        \begin{align}
            \nabla_k\Psi_T(\vf{r})
            &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
        \end{align}
        whereas the Laplacian yields
        \begin{align}
            \nabla^2_k\Psi_T(\vf{r})
            &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}),
        \end{align}
        where $d$ is the dimensionality of the problem determined by $\vf{r}_k
        \in \mathbb{R}^d$. We can thus use the gradient to find an expression
        for the drift force for particle $k$.
        \begin{align}
            \vf{F}_k(\vf{r})
            &= -4\alpha\vf{r}_k.
        \end{align}
        Using the Laplacian we can compute the kinetic term in the expression
        for the local energy. We get
        \begin{align}
            E_L(\vf{r})
            &=
            \sum_{i = 1}^N
            \Biggl(
                -\frac{\hbar^2}{2m}
                \bigl[
                    -2d\alpha + 4\alpha^2 r_i^2
                \bigr]
                + \half m\omega^2 r_i^2
            \Biggr).
        \end{align}
        In natural units, $\hbar = c = 1$, and with the mass set to unity, $m =
        1$, this reduces to
        \begin{align}
            E_L(\vf{r})
            &=
            \alpha dN
            + \biggl(
                \half\omega^2
                - 2\alpha^2
            \biggr)
            \sum_{i = 1}^N r_i^2.
            \label{eq:closed_form_natural_units_local_energy}
        \end{align}
        It is worth noting that for $\alpha = \omega/2$\footnote{Note that
        $\alpha$ is required to be positive.} we get a stable value for the
        local energy.
        \begin{align}
            E_L(\vf{r})
            &\equiv
            E_L
            =
            \frac{\omega d N}{2},
            \label{eq:minimum_energy_simple}
        \end{align}
        which is the stable minimum for $\alpha = \omega/2$.

        \subsubsection{Exact variational energy}
            As the system is non-interacting and consisting of Gaussians we can
            find an expression for the exact energy as a function of the
            variational parameter $\alpha$, i.e.,
            \begin{align}
                E(\alpha)
                &=
                \frac{\bra{\Psi_T}H\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}}
                \\
                &=
                \left(
                    \frac{\hbar^2 \alpha}{2m}
                    + \frac{m\omega^2}{8\alpha}
                \right)dN.
                \label{eq:exact_energy}
            \end{align}
            By minimizing this expression with respect to $\alpha$ yields
            \begin{align}
                \dod[]{E(\alpha)}{\alpha} = 0
                \implies
                \alpha_0 = \frac{m\omega}{2\hbar},
            \end{align}
            which in natural units reduces to $\alpha_0 = \omega/2$. The energy
            at this value of $\alpha$ (in natural units) is then
            \begin{align}
                E(\alpha_0)
                &=
                \frac{\omega dN}{2},
            \end{align}
            in perfect accordance with \autoref{eq:minimum_energy_simple} which
            yields high hopes for the latter equation and the expected minimum.

>>>>

    \subsection{Non-interacting harmonic oscillators}
        We start by looking at a simple system of non-interacting harmonic
        oscillators, where $a = 0$ and $\beta = 1$. This means that there is no
        hard-shell interaction between the particles and that we use a spherical
        potential trap.  We thus get the trial  wavefunction
        \begin{align}
            \Psi_T(\vf{r})
            = \Phi_T(\vf{r})
            = \prod_{i = 1}^N \exp\para{
                -\alpha |\vf{r}_i|^2
            }.
        \end{align}
        As $a = 0$ the interaction term, $w(\vf{r}_i, \vf{r}_j)$, vanishes and
        the Hamiltonian is given by (in the spherical case)
        \begin{align}
            H &= \sum_{i = 1}^N h(\vf{r}_i)
            = \sum_{i = 1}^N \Biggl(
                -\frac{\hbar^2}{2m}\nabla_i^2
                + \half m \omega^2 r_i^2
            \Biggr).
        \end{align}
        To find the drift force and the local energy we have to compute the
        gradient and the Laplacian of the trial wavefunction. The gradient is
        given by
        \begin{align}
            \nabla_k\Psi_T(\vf{r})
            &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
        \end{align}
        whereas the Laplacian yields
        \begin{align}
            \nabla^2_k\Psi_T(\vf{r})
            &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}).
        \end{align}
        Here $d$ is the dimensionality of the problem determined by $\vf{r}_k
        \in \mathbb{R}^d$. We can thus use the gradient to find an expression
        for the drift force for particle $k$.
        \begin{align}
            \vf{F}_k(\vf{r})
            &= -4\alpha\vf{r}_k.
        \end{align}
        Using the Laplacian we can compute the kinetic term in the expression
        for the local energy. We get
        \begin{align}
            E_L(\vf{r})
            &=
            \sum_{i = 1}^N
            \Biggl(
                -\frac{\hbar^2}{2m}
                \bigl[
                    -2d\alpha + 4\alpha^2 r_i^2
                \bigr]
                + \half m\omega^2 r_i^2
            \Biggr).
        \end{align}
        In natural units, $\hbar = c = 1$, and with the mass set to unity this
        reduces to
        \begin{align}
            E_L(\vf{r})
            &=
            \alpha dN
            + \biggl(
                \half\omega^2
                - 2\alpha^2
            \biggr)
            \sum_{i = 1}^N r_i^2.
            \label{eq:closed_form_natural_units_local_energy}
        \end{align}
        It is worth noting that for $\alpha = \omega/2$\footnote{Note that
        $\alpha$ is required to be positive.} we get a stable value for the
        local energy.
        \begin{align}
            E_L(\vf{r}; \alpha = \omega/2)
            &\equiv
            E_L
            =
            \frac{\omega d N}{2},
            \label{eq:minimum_energy_simple}
        \end{align}
        which turns out to be the exact minimum for the ground state energy of
        this system.

        \subsubsection{Exact variational energy}
            As the system is non-interacting and consisting of Gaussians we can
            find an expression for the exact energy as a function of the
            variational parameter $\alpha$, i.e.,
            \begin{align}
                E(\alpha)
                &=
                \frac{\bra{\Psi_T}H\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}}
                =
                \left(
                    \frac{\hbar^2 \alpha}{2m}
                    + \frac{m\omega^2}{8\alpha}
                \right)dN.
                \label{eq:exact_energy}
            \end{align}
            By minimizing this expression with respect to $\alpha$ yields
            \begin{align}
                \dod[]{E(\alpha)}{\alpha} = 0
                \implies
                \alpha_0 = \frac{m\omega}{2\hbar},
            \end{align}
            which in natural units reduces to $\alpha_0 = \omega/2$. The energy
            at this value of $\alpha$ (in natural units) is then
            \begin{align}
                E(\alpha_0)
                &=
                \frac{\omega dN}{2},
            \end{align}
            in perfect accordance with \autoref{eq:minimum_energy_simple} which
            yields high hopes for the latter equation and the expected minimum.
